---
title: Looking a bit closer at Ludden data
author: "Joe Lewis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

# Background

Poisson-lognormal models overpredict the number of people in which we will find
*no* strains of E coli - is this just because these data are not included in
the dataset?

Look at the Ludden paper to explore this possibility.

```{r setup}
library(tidyverse)
library(janitor)
library(lme4)

# plot model function

simulate_and_plot_from_fitted_model <- function(fitted_model,
                                                raw_data,
                                                study_name,
                                                nsubj = 1000,
                                                nsim = 10000) {
  raw_data |>
    count(n_clones) |>
    mutate(tot = sum(n)) |>
    rowwise() |>
    transmute(
      value = n_clones,
      med = n / tot,
      lci = binom.test(n, tot)$conf.int[[1]],
      uci = binom.test(n, tot)$conf.int[[2]],
      type = "raw data"
    ) -> raw_data_props

  simulate(fitted_model,
    newdata = data.frame(pid = 1:nsubj),
    allow.new.levels = TRUE,
    nsim = nsim,
    re.form = NA
  ) -> sims_df

  sims_df |>
    pivot_longer(everything()) |>
    count(name, value) |>
    group_by(name) |>
    mutate(prop = n / sum(n)) |>
    group_by(value) |>
    summarise(
      lci = quantile(prop, 0.025),
      med = median(prop),
      uci = quantile(prop, 0.975),
      type = "model fit"
    ) |>
    ggplot(aes(value, med, ymin = lci, ymax = uci, color = type, fill = type)) +
    geom_line() +
    geom_ribbon(color = NA, alpha = 0.3) +
    geom_line(data = raw_data_props) +
    geom_ribbon(data = raw_data_props, alpha = 0.3, color = NA) +
    labs(title = paste0("Study: ", study_name)) -> plot_a


  sims_df |>
    pivot_longer(everything()) |>
    ggplot(aes(value)) +
    stat_ecdf(geom = "line") +
    geom_hline(yintercept = 0.95, linetype = "dotted") +
    scale_x_continuous(breaks = c(0:10, 20, 30, 40)) +
    labs(title = paste0("Study: ", study_name)) -> plot_b

  return(list(plot_a, plot_b))
}
```

```{r load-data-and-plot}

read_csv("ludden_supp_data_S2.csv") |>
janitor::clean_names() -> ludden_df

ludden_df |>
mutate(esbl = if_else(esbl == "AmpC", "ESBL",esbl)) |>
count(patient_id, st, esbl) |>
mutate(esbl = as.factor(esbl)) |>
select(-n) |>
count(patient_id, esbl, .drop = FALSE) |>
ggplot(aes(n, fill = esbl)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ esbl) +
  labs(title = "Number of STs per participant",
  subtitle = "lumping ESBL and ampC together")

```

# Fit a model coutning ESBL strains only - including no strains found


```{r ludden_diversity_model_Title}

ludden_df |>
mutate(esbl = if_else(esbl == "AmpC", "ESBL",esbl)) |>
count(patient_id, st, esbl) |>
mutate(esbl = as.factor(esbl)) |>
select(-n) |>
rename(pid = patient_id) |>
count(pid, esbl, .drop = FALSE, name = "n_clones") |>
filter(esbl == "ESBL") -> mod_df



glmer(n_clones ~ (1 | pid), 
      family = "poisson",
      data = mod_df, 
      control = glmerControl(optimizer = "bobyqa")) -> m_lud

m_lud


simulate_and_plot_from_fitted_model(
  fitted_model = m_lud,
  raw_data = mod_df,
  study_name = "Ludden")

```

# Conclusions

- Model is a pretty good fit  
- Model is singular - probably a simple Poisson would do as well
- Strain diversity (in terms of ST) of ESBLs is not very high: in this
  situation maybe should remove all the people with 0 strains to decide on a
  number of colony picks - e.g. below, where 2 picks would do.

  ```{r ditch-0-strains}


  simulate(m_lud,
    newdata = data.frame(pid = 1:1000),
    allow.new.levels = TRUE,
    nsim = 10000,
    re.form = NA
  ) -> sims_df

  sims_df |>
  pivot_longer(everything()) |>
filter(value > 0) |>
    ggplot(aes(value)) +
    stat_ecdf(geom = "line") +
    geom_hline(yintercept = 0.95, linetype = "dotted") +
    labs(title = paste0("Study: Ludden"))

  
  ```
