---
title: Sample size calculation for Sarah's _E. coli_ diversity experiments
author: "Joe Lewis"
output: html_document
---

# Setup

The starting point is that we want to run the MGEMS/MSWEEP pipeline on a number
of stool samples alongside single colony picks and WGS. We want to quantify
diversity missed by either method and compare assemblies from WGS and MGEMS.

We assume that the number of ESBL strains within an individual is
Poisson-lognormal distributed (i.e. the number of species is given by a Poisson
distribution where the Poisson rate parameter, $\lambda$ is itself log-normally
distributed), i.e.

log($\lambda) \sim Normal(\mu, \sigma)$

The distribution is _zero-truncated_ - that is, all samples with 0 strains are
dropped, because we won't pick any strains from samples with no ESBLS; hence we
want a _zero truncated Poisson-lognormal distribution_. We assume that we'll
pick $n$ colony picks from each sample where any ESBL is identified, where
previous fits to data from the literature suggests $n=6$ colony picks has a
good chance of identifying all strains (where, importantly, strain is defined
at the ST level or some other high level clustering method) in a sample.

How should we decide how many samples to use to compare MSWEEP/MGEMS to WGS
colony pick strategies? On a approach would be to say that we want to know in
what proportion of samples we miss one or more ESBL strains with WGS; if we are
interested in transmission, then this proportion is possible missed
transmission events. We can power a comparison to define this proportion to a
certain accuracy; say +/- 10% (an arbitrary choice!). This will depends on
$\mu$, $\sigma$, and $n$; we can estimate values of $\mu$ and $\sigma$ from the
fits we have made to the data.

The sample size calculation method will therefore be:

- Pick some likely values for $\mu$ and $\sigma$ based on available data
- For these values, use the cumulative distribution function of the
  zero-truncated Poisson lognormal distribution to determine the proportion of
  samples that will have > $n$ strains
- Calculate the number of samples that will be needed to estimate this
  proportion with an accuracy +/10%


# Zero-truncated poisson-lognormal distributions

First plot the distributions for a variety of parameter values. The range of
$\mu$ from fitting to data was 0.4-1 and the range of $\sigma$ was 0-0.4. 

```{r poisson-lognormal, warning = FALSE, message = FALSE}

library(ztpln) # this package gives the zero truncated Poisson- lognormal distribution
library(tidyverse)
library(presize) # this gives function to calculate sample size

mu <- c(0.5,1,1.5)
sig <- c(0.01, 0.3, 0.6)


# first plot the distributions for given parameter values

crossing(mu,sig, x = 1:10) |>
rowwise() |>
mutate(dens = dztpln(x = x, mu = mu, sig = sig),
       mu = paste0("mu = ", mu),
       sig = paste0("sigma =", sig)) |>
ggplot(aes(x,dens)) +
  geom_col() +
  facet_grid(mu ~ sig)

```

# Sample size calculations

I've added a black horizontal line at 25 samples - for the parameter values
that fit the data we have ($\mu$ up to 1 and $\sigma$ up to 0.3) this will
allow estimated on proportion of samples with missed strains +/- 10%.

This would involve sequencing of 25x6 = 150 colonies, which seems reasonable. I
think there is advantage in picking a small number of samples/ colony picks
initially - if we find that we need to do more, we can always do that!

```{r sim-studies, message = FALSE, warning = FALSE}

# define ranges of parameter values

mu <- seq(from = 0.3, to = 1.5, by = 0.1)
sig <- c(0.01, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6)

# define a cumulative probability function
pztpln <- function(n,mu,sig) {
  sum(dztpln(1:n,mu,sig))
}


#calculate and plot

crossing(mu, sig,
  cutoff = 6
) |>
  rowwise() |>
  mutate(
    missed_prop = 1 - pztpln(cutoff, mu, sig),
    n = prec_prop(p = missed_prop, conf.width = 0.2, method = "wilson" )$n,
    sig = as.character(sig)
  ) |>
  ggplot(aes(mu, n, group = sig, color = sig)) +
  geom_line() +
  facet_wrap(~cutoff) +
  labs(
    color = "sigma",
    title = "n to estimate proportion samples with missed strains +/- 10%"
  ) +
  geom_hline(aes(yintercept = 25))


```
